---
output: github_document
---
<!-- README.md is generated from README.Rmd. Please edit that file -->
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# bmm   <!-- badges: start -->
  [![R-CMD-check](https://github.com/venpopov/bmm/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/venpopov/bmm/actions/workflows/R-CMD-check.yaml)
  [![test-coverage](https://github.com/venpopov/bmm/actions/workflows/test-coverage.yaml/badge.svg)](https://github.com/venpopov/bmm/actions/workflows/test-coverage.yaml)
<!-- badges: end -->

The goal of bmm (Bayesian Measurement Models) is to make it easier to estimate 
common measurement models for behavioral research using Bayesian hierarhical 
estimation via the 'brms' package'. Currently implemented models are:

```{r, results="asis", echo=FALSE}
bmm::print_pretty_models_md()
```

You can always view the latest list of supported models by running:

```{r}
bmm::supported_models()
```

## Installation 

You can install the development version of bmm from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("venpopov/bmm")
```

If you want to install the package together with the vignettes, you can use:

``` r
devtools::install_github("venpopov/bmm", build_vignettes = TRUE)
```

The package was significantly updated on Feb 03, 2024. If you are following the old version of the [Tutorial preprint](https://osf.io/preprints/psyarxiv/umt57), you need to install the 0.0.1 version of the bmm package with:

``` r
# install.packages("devtools")
devtools::install_github("venpopov/bmm@v0.0.1")
```

## Example 1

The three-parameter mixture model by Bays et al (2009) assumes that
responses can come from three different sources - noisy representation of the
target, confusion with noisy representation of non-target items, or guessing
based on a uniform distribution. To estimate these parameters for a dataset, we
can use the `fit_model()` function. First, let's generate a dataset with known
parameters. We can use the function `gen_3p_data()`

```{r example, message=FALSE, warning=FALSE}
library(bmm)
library(tidyverse)
dat <- gen_3p_data(N=2000, pmem=0.6, pnt=0.3, kappa=10, setsize=4, relative_resp=T)
head(dat)
```

We have a dataset of 2000 observations of response error, of which 60%
(pmem=0.6) come from the target distribution, 30% (pnt=0.3) are non-target
swaps, and 10% are guessing. The precision of the von Mises distribution is 10,
the presented setsize is 4 (one target and three lures/non-targets), and the
values are coded relative to the target value (i.e., response error for the y
variable or displacement relative to the target for the lures).

Just for visualization purposes, here's a histogram of the error distribution,
demonstrating a typical pattern - a normal distribution centered on 0, with long
tails:

```{r}
hist(dat$y, breaks = 60, xlab = "Response error relative to target")
```
Another key property of the data is that some error responses are not random,
but that they are due to confusion of the target with one of the lures. We can
visualize this by centering the response error relative to each of the possible
non-target locations. We do this with the helper function
`calc_error_relative_to_nontargets()`:

```{r}
dat %>% 
  calc_error_relative_to_nontargets('y', paste0('nt',1:3,'_loc')) %>% 
  ggplot(aes(y_nt)) +
  geom_histogram()
```

Ok, so now let's fit the three-parameter model. We only need to do two things:
- Specify the model formula
- Call fit_model()

In this example the parameters don't vary over conditions, so we have no
predictors. `y` is the name of the response error variable, whereas `kappa`,
`thetat` and `thetant` are the parameters of the model - precision, mixing
proportion for correct responses and mixing proportion for non-target swaps.

```{r}
ff <- brms::bf(y ~ 1,
         kappa ~ 1,
         thetat ~ 1,
         thetant ~ 1)
```

Then specify the model and give it information about the required arguments. In
the case of the 3-parameter model, we need to specify the names of the non-target
variables and the setsize. We can do this with the `mixture3p()` function:


```{r}
model <- mixture3p(non_targets = paste0('nt',1:3,'_loc'), setsize=4)
```

You can always get full help and information about the model and its required 
arguments, as well as examples by running `?mixture3p`

Finally we just run the model. The arguments to the function explained in
`help(fit_model)` and you can also pass any additional arguments that you would
pass to `brm`.

``` r
fit <- fit_model(formula = ff,
                 data = dat,
                 model = model,
                 parallel=T,
                 iter=500,
                 backend='cmdstanr')
```

