---
title: "Introduction to `bmmformula` syntax"
output: bookdown::html_document2
author: 
  - Ven Popov
  - Gidon Frischkorn
bibliography: REFERENCES.bib
header-includes:
  - \usepackage{amsmath}
vignette: >
  %\VignetteIndexEntry{Intorduction: bmmformula}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
pkgdown:
  as_is: true
---
<style type="text/css">
div.main-container {
max-width: 800px !important;
}

p {
margin-top: 1.5em ;
margin-bottom: 1.5em ;
}
.author{
    display: none;
}
</style>

```{r, include = FALSE}
options(crayon.enabled = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "jpeg",
  dpi = 100
)
fansi::set_knit_hooks(knitr::knit_hooks, which = c("output","message","error"))

pacman::p_load(bmm)
```

# The `bmmformula` syntax

The `bmmformula` syntax generally follows the same principles as the formula syntax for [`brms` models](https://rdrr.io/cran/brms/man/brmsformula.html). A `bmmformula` follows the general structure:

```r
parameter ~ peffects + ( geffects | group)
```

However, instead of predicting a `response` - as is typically done in `brms` - in `bmm` a `bmmformula` predicts a `parameter` from a `bmmmodel`[^1]. The `peffects` part specifies all effects that are assumed to be the same across observations. In typical vocabulary such effects are called 'population-level' or 'overall' effects. The `geffects` part specifies effects that are assumed to vary across grouping variables specified in `group`. Typically, these effects are called 'group-level' or 'varying' effects.

[^1]: You can always find information on the parameters of a `bmmmodel` by calling the help for the model you want to fit: `?bmmmodel`

As you will typically, provide a `bmmformula` for each parameter of a `bmmmodel` we recommend to set up the formula in a separate object before passing it to the `fit_model` function. There are two ways to set up formulas, both work equally easy and well:

1. a single call to the function `bmmformula` or its short form `bmf` with separating formulas for different parameters by comma. 
2. add `bmmformulas` for each parameters to an object by using the `+` operator. 

Ηere are two examples for each method that result in the same `bmmformula` object:

``` {r}
my_formula <- bmmformula(
  thetat ~ 1 + set_size,
  thetant ~ 1 + set_size,
  kappa ~ 1
)

my_formula <- bmf(thetat ~ 1 + set_size) +
  bmf(thetant ~ 1 + set_size) +
  bmf(kappa ~ 1)
  
my_formula
```

As `bmmformula` syntax builds upon `brmsformula` syntax, you can use any functionality you might use in `brms` for specifying a `bmmformula`. One difference is, that you _do not_ have to explicitly specify if a formula is a non-linear formula, as `bmm` recognizes a formula as non-linear as soon as one of the left-hand side arguments of a formula is also used as a right-hand side argument in the whole formula. So the argumetn `nl` does not exist for `bmmformula`. Similarly, adding grouping statements around group-level effects to estimate them separately for different sub-groups in your design can be added. So for example, a more complicated `bmmformula` could look like this:

```r
complex_formula <- bmf(
  thetat ~ end_theta + (start_theta - end_theta) * exp(-R*ss_num),
  # other model parameters
  thetant ~ 1 + (1 | gr(ID, by = age_group)),
  kappa ~ 1 + (1 | gr(ID, by = age_group)),
  # parameters of the non-linear function
  start_theta ~ 1 + (1 | gr(ID, by = age_group)),
  end_theta ~ 1 + (1 | gr(ID, by = age_group)),
  R ~ 1 + (1 | gr(ID, by = age_group))
)
```

This formula implements an exponential reduction in `thetat` from a `start_theta` at `ss_num = 0` towards the lower asymptote `end_theta` that will be reached with the slope `R`. In addition, the group level effects have been grouped by `age_group` so that the variation in the `Intercept` would be estimated separately for different levels of the `age_group` variable. This formula just serves as an example how `bmmformula` syntax allows to use the complex formula syntax implemented in `brms` and passes the information on to `brms` without losing any of the functionality. 

# Seperating response variables from model parameters

As the response is not explicit from the `bmmformula`, in `bmm` the response variables, and also other variables relevant for a `bmmmodel`, are linked to the model when setting up the `bmmmodel` object. For example, when fitting the `mixture3p` model, you have to provide the names of the variables that reflect the response error (e.g., `y`), the location of the non-target features (e.g. `nt_col1`, ..., `nt_col4`), and the set-size (e.g., `ss`) in our data:

```r
my_model <- mixture3p(resp_err = "y", 
                     nt_features = paste0("nt_col",1:4), 
                     set_size = "ss")
```

Internally, `bmm` will then link the response variables to the `bmmformula` given for the parameters of a `bmmmodel`.

# Fixing parameters to constant values

In some cases it can be reasonable to fix parameters to a specific value. This can either be done via the `bmmformula` or via a [constant prior](https://github.com/paul-buerkner/brms/issues/783#issuecomment-577909409) specified for the parameter to be fixed. Here we will only outline fixing parameters via the `bmmformula`.

To fix a parameter via the `bmmformula` you have to set the parameter to the value it should be fixed to by using `=` instead of `~`. For example, the following formula would fix `kappa` to the value `1`.

```r
fixed_formula <- bmf(kappa = 1)
```

When fixing a parameter to a constant value, you need to keep in mind that the value you fix it to, is the value on the parameter scale. Oftentimes, `bmm` uses link functions to convert parameters from a bounded native scale (e.g. only positive values for precision parameters, such as `kappa`) to an unbounded parameter scale. In the case of `kappa`, `bmm` uses a `log` link function. Thus, when fixing the parameter to `1`, this will effectively fix the parameter to $e^1$ on the native scale.

# Accessing the `brmsformula` generated by `bmm`

```{r, include=FALSE, warning=FALSE, message=FALSE}
bmmfit <- bmm:::bmmfit_mixture2p_vignette
```

Let's assume you have fit a `mixture2p` model to a data set `my_data` as in the `vignette("mixture_models")`. For this, you had to specify a `bmmformula` and model object:

```r
user_formula <- bmf(
  thetat ~ 0 + set_size + (0 + set_size | id),
  kappa ~ 0 + set_size + (0 + set_size | id)
)

my_model <- mixture2p(resp_err = "error")

bmmfit <- fit_model(
  formula = user_formula,
  data = my_data,
  model = my_model,
)
```

If you were now interested to see how the `bmmformula` that you passed to `bmm` is converted to a `brmsformula`, you can access the `brmsformula` via the `bmmfit` object that will be returned by the `fit_model` function. Accessing the `formula` object stored in the fit object via `bmmfit$formula` will print out the `brmsformula` generated by `bmm` for the specified model. Thus, you can access the `brmsformula` generated for this model by calling `bmmfit$formula`:

```{r}
bmmfit$formula
```

Similarly, you can access the distributional family that was used to implement the specified model, so `bmmfit$family` will return the family object that was generated by `bmm` and then passed to `brms` for fitting. 

```{r}
bmmfit$family
```

Finally, you can also access the data that was used by `bmm` to fit the model via `bmmfit$data`. In many cases this data will be equal to the data provided by the user. But sometimes `bmm` internally computes additional index variables for specifying the models adequately. The data stored in the `bmmfit` object contains these additional variables.

```{r}
head(bmmfit$data)
```

This way, should you be interested to customize models to fit them in `brms` without `bmm`, you are able to obtain the most important information that is essential for specifying the models implemented in `bmm`. 

Fit objects from `bmm` use a custom summary function to format the output. However, you can call a `brms` summary instead of a `bmm` summary by setting the `backend` option in the summary function to `brms`. This way the summary function for `brmsfit` objects will be used instead of the function for `bmmfit` objects. The `brmsfit` summary also contains the `brmsformula` that is created by `bmm`.

``` {r}
summary(bmmfit, backend = "brms")
```
