---
title: "The Signal Discrimination Model (SDM)"
output: bookdown::html_document2
author: 
  - Ven Popov
  - Gidon Frischkorn
bibliography: REFERENCES.bib
header-includes:
  - \usepackage{amsmath}
vignette: >
  %\VignetteIndexEntry{The Signal Discrimination Model (SDM)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
pkgdown:
  as_is: true
---
<style type="text/css">
div.main-container {
max-width: 800px !important;
}

p {
margin-top: 1.5em ;
margin-bottom: 1.5em ;
}
.author{
    display: none;
}
</style>

```{r, include = FALSE}
options(crayon.enabled = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "jpeg",
  dpi = 100
)
fansi::set_knit_hooks(knitr::knit_hooks, which = c("output","message","error"))
```

# Introduction to the model

The Signal Discrimination Model is a measurement model for continuous
reproduction tasks in the visual working memory domain. The model was originally
introduced by @Oberauer_2023. As other measurement models for continuous
reproduction tasks, its goal is to model the distribution of angular response
errors.

The model assumes that when a test probe appears, all possible responses on the
circle ($\theta$) are activated with a strength that depends on the distance
between the feature stored in memory ($\mu$) and the response options. Formally,
this is given by the following activation function:


$$
S(\theta) = c \cdot \frac{\exp(\kappa \cdot \cos(\theta-\mu))}{2\pi I_0(\kappa)}
$$


where $c$ is the memory strength parameter, $\kappa$ is the precision parameter,
and $I_0$ is the modified Bessel function of the first kind of order 0. Thus,
the activation function follows a von Mises distribution, weigthed by a memory
strength parameter.

The activation of response options is corrupted by noise, which is assumed to
follow a Gumbel distribution. Then the response is the option with the highest
activation value:

$$
Pr(\theta) = argmax(S(\theta) + \epsilon) \\ \epsilon \sim Gumbel(0,1)
$$
This is equivalent to the following softmax function (also known as the
exponentiated Luce's choice rule):

$$
Pr(\theta) = \frac{\exp(S(\theta)}{\sum_{i=1}^{n} \exp(S(\theta_i))}
$$

where n is the number of response options, most often 360 in typical visual
working memory experiments.

In summary, the model assumes that response errors come from the following
distribution, where $\mu = 0$:

$$
\Large{f(\theta\ |\ \mu,c,\kappa) = \frac{e^ {c \ \frac{e^{k\ cos(\theta-\mu)}}{2\pi I_o(k)}}}{Z}}
$$

and Z is the normalizing constant to ensure that the probability mass sums to 1.

# Parametrization in the `bmm` package

In the `bmm` package we use a different parametrization. The parametrization is
chosen for numerical stability and efficiency. Three features of the
parametrization above make it difficult to work with in practice. First, the
modified bessel function $I_0$ increases rapidly, often leading to numerical
overflow. Second, the bessel function is expensive to compute, and estimating
this model with MCMC methods can be slow. Third, the normalizing constant in the
denominator requires summing 360 terms, which is also slow.

To address these issues, we use the following parametrization of the SDM distribution:

$$
\Large{f(\theta\ |\ \mu,c,\kappa) = \frac{
e^{c \ \sqrt{\frac{k}{2\pi}} e^{k \ (cos(\theta-\mu)-1)}}
}{Z}}
$$

This parametrization is derived from the known approximation of the modified
bessel function for large $k$ (@abramowitz1988handbook):

$$
I_0(\kappa) \sim ~ \frac{e^{\kappa}}{\sqrt{2\pi \kappa}}, \ \ \ \ \kappa \rightarrow \infty
$$

If needed, the $c$ parameter of the original formulation by Oberauer (2023) can
be computed by:

$$
c_{oberauer} = c_{bmm} \ e^{-\kappa} I_0(\kappa)\sqrt{2 \pi \kappa}
$$

This parametrization does not change the predicted shape of the distribution,
but it produces slightly different values of $c$ for small values of $kappa$.
The parametrization is the default in the `bmm` package.

A second optimization concerns the calculation of the normalizing constant $Z$.
The original model assumed that responses can only be one of 360 discrete
values, resulting in a probability mass function. In `bmm` we treat the response
variable as continuous, which makes $f(\theta)$ a probability density function.
This means that we can calculate the normalizing constant $Z$ by integrating
$f(\theta)$ over the entire circle:

$$
Z = \int_{-\pi}^{\pi} f(\theta) d\theta
$$

This integral cannot be expressed in closed form, but it can be approximated
using numerical integration methods. The results between the discrete and
continuous formulations are nearly identical, for large number of response
options (as in typical applications), but not when the number of response
options is small, for example in 4-AFC tasks.

# Fitting the model with `bmm`

Begin by loading the `bmm` package:

```{r setup, message=FALSE, warning=FALSE}
library(bmm)
```

## Generating simulated data

If you already have data you want to fit, you can skip this section. For the
current illustration, we will generate some simulated data with known
parameters. `bmm` provides density functions in typical R style, with the prefix
`d` for density (`dsdm`), `p` for the cumulative distribution function (`psdm`),
`q` for the quantile function (`qsdm`), and (`rsdm`) for generating random
deviates. Here's how to simulate data from the SDM distribution for three
conditions:


```{r}
# set seed for reproducibility
set.seed(123)

# define parameters:
cs <- c(2, 9, 2)
kappas <- c(3, 1, 8)

# simulate data from the model
y <- c(rsdm(n = 1000, mu=0, c = cs[1], kappa = kappas[1], parametrization = "sqrtexp"),
       rsdm(n = 1000, mu=0, c = cs[2], kappa = kappas[2], parametrization = "sqrtexp"),
       rsdm(n = 1000, mu=0, c = cs[3], kappa = kappas[3], parametrization = "sqrtexp"))
dat <- data.frame(y = y, cond = factor(rep(c('A','B','C'), each=1000)))
```

This gives us the following distribution of response errors, with lines
overlaying the predicted density generated with `dsdm`:

```{r warning=FALSE, message=FALSE, fig.width=8, fig.height=3}
# generate predicted SDM density:
dd <- data.frame(y = rep(seq(-pi, pi, length.out=1000),3),
                 cond = factor(rep(c('A','B','C'), each=1000)),
                 c = rep(cs, each=1000),
                 kappa = rep(kappas, each=1000))
dd$d <- dsdm(dd$y, mu = 0, c = dd$c, kappa = dd$kappa, parametrization = "sqrtexp")

# prepare labels for plots
par_labels <- data.frame(cond = c('A','B','C'),
                         c = c(2, 9, 2),
                         kappa = c(3, 1, 8))
par_labels$label <- paste0('c = ', par_labels$c, '\nkappa = ', par_labels$kappa)

# plot the data and the predicted density
library(ggplot2)
ggplot(dat, aes(x=y, fill=cond)) +
  geom_histogram(aes(y=..density..), binwidth = 0.1, position = "identity", alpha = 0.5) +
  theme_classic() +
  facet_wrap(~cond) +
  geom_line(data=dd, fun=dsdm, aes(y, d)) +
  scale_x_continuous(limits = c(-pi, pi)) +
  labs(title = "Simulated data from the SDM distribution",
       x = "Response error (radians)",
       y = "Density") +
  geom_text(data=par_labels, aes(x = -pi, y = 0.5, label = label), 
            hjust = 0, vjust = 0, size = 3)
```

## Estimating the model with `bmm`

To estimate the parameters of the SDM distribution, we can use the `bmm()`
function. First, let's specify the model formula. We want `c` and `kappa` to
vary over conditions. The first two lines of the formula specify how the
parameters vary over conditions. In this case, we want them to vary over the
`cond` variable, so we use `c ~ 0 + cond` and `kappa ~ 0 + cond`:

```{r}
ff <- bmf(c ~ 0 + cond, kappa ~ 0 + cond)
```

Then we specify the model, which in this case is just `sdmSimple()`, and we
provide the name of the response error variable in the dataset:

```{r}
model <- sdm(resp_error = "y")
```

Finally, we can fit the model. We strongly recommend using `cmdstanr` as a
backend for fitting the SDM model, as it is much faster and more stable than the
default `rstan` backend for this particular model. Here's how to fit the model
with `cmdstanr`:


```{r}
fit <- bmm(
  formula = ff,
  data = dat,
  model = model,
  cores = 4,
  init = 0.5,
  backend = 'cmdstanr',
  file = "assets/bmmfit_sdm_vignette.rds"
)
```

The model takes about 30 seconds to fit after it is compiled. 

We can now inspect the results of the model fit:

```{r, comment=NA, collapse=TRUE}
summary(fit)
```

We see that all Rhat values are less than 1.01, which is a good sign that the
chains have converged. In principle you have to do more inspection, but let us
see the estimated parameters. The model uses a log-link function for the `c` and
`kappa` parameters, so we have to exponentiate the coefficients to get the
estimated parameters:

```{r}
exp(brms::fixef(fit)[2:7,1])
```

which are close to the true values we used to simulate the data:

```{r}
par_labels[,1:3]
```

We can see that even though the estimated parameters are close, they are not
exactly the same as the true parameters. To get a better picture, we can plot
the estimated posterior distributions of the parameters:

```{r, message=FALSE, warning=FALSE, fig.width=5}
library(tidybayes)
library(dplyr)
library(tidyr)

# extract the posterior draws
draws <- tidybayes::tidy_draws(fit)
draws <- select(draws, b_c_condA:b_kappa_condC)

# plot posterior with original parameters overlayed as diamonds
as.data.frame(draws) %>% 
  gather(par, value) %>% 
  mutate(value = exp(value)) %>% 
  ggplot(aes(value, par)) +
  tidybayes::stat_halfeyeh(normalize = "groups") +
  geom_point(data = data.frame(par = c('b_c_condA', 'b_c_condB', 'b_c_condC',
                                       'b_kappa_condA', 'b_kappa_condB', 'b_kappa_condC'),
                               value = c(cs, kappas)),
             aes(value, par), color = "red",
             shape = "diamond", size = 2.5) +
  scale_x_continuous(lim = c(0, 20))
```

The true parameters all lie within the 50% credible intervals, which is a good
sign that the model is able to recover the true parameters from the data.

As a final step, we can plot our data again, adding another line overlay with
the density predicted by the estimated parameters:

```{r warning=FALSE, message=FALSE, fig.width=8, fig.height=4}
# generate predicted SDM density:
ddest <- data.frame(
  y = rep(seq(-pi, pi, length.out = 1000), 3),
  cond = factor(rep(c("A", "B", "C"), each = 1000)),
  c = rep(exp(brms::fixef(fit)[2:4, 1]), each = 1000),
  kappa = rep(exp(brms::fixef(fit)[5:7, 1]), each = 1000)
)

# prepare labels for plots
par_labels <- data.frame(
  cond = c("A", "B", "C"),
  c_est = exp(brms::fixef(fit)[2:4, 1]),
  kappa_est = exp(brms::fixef(fit)[5:7, 1]),
  c = c(2, 9, 2),
  kappa = c(3, 1, 8)
)
par_labels$label <- paste0(
  "c = ", par_labels$c, " (estimate = ", round(par_labels$c_est, 2), ")\n",
  "kappa = ", par_labels$kappa, " (estimate = ", round(par_labels$kappa_est, 2), ")"
)

ddest$d <- dsdm(ddest$y, mu = 0, c = ddest$c, kappa = ddest$kappa, parametrization = "sqrtexp")

# plot the data and the predicted density
ggplot(dat, aes(x = y, fill = cond)) +
  geom_histogram(aes(y = ..density..), binwidth = 0.1, position = "identity", alpha = 0.5) +
  theme_classic() +
  facet_wrap(~cond) +
  geom_line(data = dd, fun = dsdm, aes(y, d), color = "black") +
  geom_line(data = ddest, fun = dsdm, aes(y, d), color = "red") +
  scale_x_continuous(limits = c(-pi, pi)) +
  labs(
    title = "Simulated data from the SDM distribution",
    x = "Response error (radians)",
    y = "Density"
  ) +
  geom_text(data = par_labels, aes(x = -pi, y = 0.9, label = label), 
            hjust = 0, vjust = 0, size = 3) +
  scale_y_continuous(limits = c(0, 1))
```

The histograms represent the data, the black lines represent the predicted
density from the true parameters, and the red lines represent the predicted
density from the estimated parameters. We can see that the estimated parameters
are able to capture the main features of the data.

# References
